# Variance structures

# Goals

- Gain familiarity with graphical model checking
- Learn about variance structures in the nlme package
- Learn to identify and deal with violations of assumptions about variance

# A simulated data set

I've simulated a data set with the predictor `x`, the response variables `y`
through `y4`, collected for groups (`group`) A through L, and collected over
time (`time`). Let's load that data set:

```{r}
library(tidyverse)
library(nlme)
library(lme4)
d <- readRDS("data/generated/model-checking.rds")
d
# View(d)
```

All of the response variables are continuous with both negative and positive
values and can be fitted with a linear mixed effects model.

Let's quickly plot the data.

```{r}
ggplot(d, aes(x, y)) + geom_point() + facet_wrap(~group)
```

The response variable `y` can be fit without any issues with the following
model:

```{r}
m <- lme(y ~ x, random = ~ 1 | group, data = d)
summary(m)
```

Some of the most important ways that we can check our model are to plot our
fitted (predicted) values overlaid on the data, plot the residuals against the
fitted values, plot the fitted values against the observed values, and plot
residuals against each of our predictors in the model and any predictors
that were not included in the model.

What are the model assumptions that these various plots help us check? What are
we looking for when we make these plots?

We will plot the residuals various ways now using the augment function from the broom package.

```{r}
aug <- broom::augment(m)

ggplot(aug, aes(x, y)) + geom_point() + facet_wrap(~group) +
  geom_line(aes(x, .fitted), colour = "red")

ggplot(aug, aes(.fitted, .resid)) + geom_point() + facet_wrap(~group) +
  geom_hline(yintercept = 0)

ggplot(aug, aes(.fitted, y)) + geom_point() + facet_wrap(~group) +
  geom_abline(intercept = 0, slope = 1)

ggplot(aug, aes(time, .resid)) + geom_point() + facet_wrap(~group) +
  geom_hline(yintercept = 0)

ggplot(aug, aes(x, .resid)) + geom_point() + facet_wrap(~group) +
  geom_hline(yintercept = 0)
```

Alternatively, as we have done before, we can use the plot functions built into
the nlme (or the lme4) package. We will use this shortcut here and below for
speed and to quickly access the "normalized" residuals which incorporate any
variance or correlation structure.

```{r}
plot(m, resid(., type = "normalized") ~ fitted(.), abline = 0)
plot(m, resid(., type = "normalized") ~ fitted(.) | group, abline = 0)
plot(m, resid(., type = "normalized") ~ time | group, abline = 0)
plot(ACF(m, resType = "normalized"))
```

We can also check the normality of the random effects, although we don't need
to get too caught up checking these and there is no need to check these
for the rest of these exercises.

```{r}
qqnorm(ranef(m)[,"(Intercept)"])
```

# When things go wrong

There are at least 3 major ways I can think of that things might look wrong with residual plots.

1. There might be large-scale trends in the residuals.
2. There might be localized clumping or patterns in the residuals.
3. The spread of the residuals may not be constant.

## Challenge 0

Turn to your neighbor and discuss the approaches you would try to solve the first 2 problems.

# Variance structures

The third problem, a problem with the variance of the residuals, can be solved by adding covariates to the variance of the residuals.

The package nlme makes this relatively easy. (There is no other major R package that implements variance structures to my knowledge.)

There are a number of possible variance structures. You can see a list of them with `?nlme::varClasses`. The best reference on these variance structures is in Pinheiro and Bates (2000) "Mixed-Effects Models in S and S-PLUS".

The most common ones I've seen used are `nlme::varIdent`, `nlme::varExp`, and `nlme::varPower`.

Let's take a look at the shape of the exponential and power variance functions in nlme:

```{r}
get_varPower <- function(v, t1) abs(v)^(2 * t1)
get_varConstPower <- function(v, t1, t2) (t1 + abs(v)^t2)^2
get_varExp <- function(v, t1) exp(2 * t1 * v)

library(manipulate)

v <- seq(0.05, 4, length.out = 100)

manipulate({plot(v, get_varPower(v, t1), type = "l", main = "nlme::varPower()")},
  t1 = slider(-1, 2, 0.1, step = 0.1))

manipulate({plot(v, get_varConstPower(v, t1, t2), type = "l",
  main = "nlme::varConstPower()")},
  t1 = slider(-4, 4, 1, step = 0.1),
  t2 = slider(-1, 2, 0.1, step = 0.1))

manipulate({plot(v, get_varExp(v, t1), type = "l", main = "nlme::varExp()")},
  t1 = slider(-0.5, 1, 0.1, step = 0.1))
```

We can add one of these variant structures by passing it to the `weights` argument in `nlme::lme` or `nlme::gls`. E.g. `weights = varExp()`. Read the help for each of these functions to see what the default value for the argument `form` is inside the variance function. `?nlme::varExp`

# Model checking exercises

In the following challenges, there is something wrong in each case. The
solution involves adding one of `varExp`, `varIdent`, or `corAR1` temporal
correlation.

I've started each one off. You will need to plot the residuals and figure out
what variance or correlation structure is missing.

Remember that you can use the `update` function to reduce copying and pasting
of code.

# Challenge 1

```{r}
ggplot(d, aes(x, y1)) + geom_point() + facet_wrap(~group)
m1 <- lme(y1 ~ x, random = ~ 1 | group, data = d)

m1_right <- update(m1, weights = varExp(form = ~ fitted(.))) # exercise
```

# Challenge 2

```{r}
ggplot(d, aes(x, y2)) + geom_point() + facet_wrap(~group)
m2 <- lme(y2 ~ x, random = ~ 1 | group, data = d)

m2_right <- update(m2, correlation = corAR1()) # exercise
```

# Challenge 3

```{r}
ggplot(d, aes(x, y3)) + geom_point() + facet_wrap(~group)
m3 <- lme(y3 ~ x, random = ~ 1 | group, data = d)

m3_right <- update(m3, correlation = varIdent()) # exercise
```

# Challenge 4

This bonus question is harder. Hint: you can combine multiple variance structures with `varComb` and will need to do that here.

```{r}
ggplot(d, aes(x, y4)) + geom_point() + facet_wrap(~group)
m4 <- lme(y4 ~ x, random = ~ 1 | group, data = d)

m4_right <- update(m4, weights = varComb(varIdent(), varExp())) # exercise
```
