
demo again - can get a lot of zeros

hurdle, zero-inflation parameter

what if continuous with zeros? link to blog

Let's try fitting models with a zero inflation parameter.

```{r}
library(tidyverse)
library(glmmTMB)
```

Zero-inflation: With probability 1-pz, Y comes from a Poisson (or negative binomial) distribution, and with probability pz, Y is zero (Bohning et al. 1999).

Bohning, D., E. Dietz, and P. Schlattmann. 1999. The zero-inflated Poisson model and the decayed, missing and filled teeth index in dental epidemiology. Journal of the Royal Statistical Society. Series A (Statistics in Society) 162:195–209.

<https://groups.nceas.ucsb.edu/non-linear-modeling/projects/owls/WRITEUP/owls.pdf>

BB: <https://stat.ethz.ch/pipermail/r-sig-mixed-models/2012q3/018874.html>
```{r}
# simulated zero-inflated Poisson example
set.seed(99)
d <- data.frame(f = factor(rep(letters[1:10], each = 10)), x = runif(100))
u <- rnorm(10, sd = 2)
d$eta <- with(d, u[f] + 1 + 4 * x)
pz <- 0.2 # probability of excess zeros
zi <- rbinom(100, size = 1, prob = pz)
d$y <- ifelse(zi, 0, rpois(100, lambda = exp(d$eta)))
d$zi <- zi

ggplot(d, aes(x, y, colour = f, shape = as.factor(zi))) + geom_point() +
  scale_shape_manual(values = c(20, 21))

ggplot(d, aes(x, y, colour = f)) + geom_point() +
  facet_wrap(~zi)
```

```{r, eval=FALSE}
zip_admb <- glmmADMB::glmmadmb(y ~ x + (1 | f), 
  data = d, family = "poisson", zeroInflation = TRUE)
summary(zip_admb)
```

```{r}
zip_tmb <- glmmTMB(y ~ x + (1 | f), data = d, family = "poisson", ziformula = ~ 1)
summary(zip_tmb)
```

```{r}
# nb_tmb <- glmmTMB(y ~ x + (1 | f), data = d, family = "nbinom2")
# summary(nb_tmb)
```


```{r}
zi <- coef(summary(zip_tmb))$zi[1, "Estimate"]
zi_se <- coef(summary(zip_tmb))$zi[1, "Std. Error"]

# CIs:
plogis(zi - 2 * zi_se) %>% round(2)
plogis(zi) %>% round(2)
plogis(zi + 2 * zi_se) %>% round(2)
```

Hurdle / delta model:

```{r}
hurd_a <- glmmTMB(y ~ x + (1 | f), data = dplyr::filter(d, y > 0), 
  family = list(family = "truncated_poisson", link = "log"))
summary(hurd_a)
```

```{r}
d <- mutate(d, y1 = y > 0)
hurd_b <- glmmTMB(y1 ~ x + (1 | f), data = d, 
  family = binomial(link = "logit"))
summary(hurd_b)
```

Predictions:

```{r}
hurd_b_lme4 <- lme4::glmer(y1 ~ x + (1 | f), data = d, 
  family = binomial(link = "logit"))
summary(hurd_b_lme4)

pred_a <- predict(hurd_a, newdata = d)
pred_b <- predict(hurd_b_lme4, newdata = d) %>% plogis

head(pred_a)
head(pred_b)
d$pred <- pred_a * pred_b

filter(d, zi == 1 | !y1) %>%
ggplot(aes(y, pred, colour = f, shape = as.factor(zi))) + 
  geom_point() + scale_shape_manual(values = c(20, 21))

```

```{r}
AIC(zip_tmb)
AIC(hurd_a) + AIC(hurd_b)
AIC(nb_tmb)
```
