# Zero-inflated models 

# Goals 

- Gain familiarity with the concept of zero-inflated models
- Understand the difference between estimating a zero-inflated parameter and a hurdle/delta model
- Learn how to approach continuous positive data that also has zeros

# Introduction

In my experience, everyone thinks they need a zero inflated model when they work with count data but they rarely do. For example, a negative binomial distribution with a small mean can generate a ton of zeros.

For example:

```{r}
hist(rnbinom(200, size = 0.1, mu = 0.5))
```

Nonetheless, sometimes there is an issue with an excess of zeros compared to what the Poisson or negative binomial allows.

What are some situations that could cause this? 

# Adding a zero-inflation parameter 

Let's try fitting models with a zero-inflation parameter. To keep this example small we will work with a simulated data set.

If our zero inflation parameter is `pz`, then our model assumes a probability of `1-pz` that the response data comes from a Poisson (or negative binomial) distribution, and a probability of `pz` that the data is an excess zero.

To avoid reinventing the wheel we will work with a modified version of a small simulation written by Ben Bolker <https://stat.ethz.ch/pipermail/r-sig-mixed-models/2012q3/018874.html>

```{r}
library(tidyverse)
inverse_logit <- function(x) plogis(x)
```


```{r}
set.seed(123)
d <- data.frame(f = factor(rep(letters[1:10], each = 10)), x = runif(100))
u <- rnorm(10, sd = 2)
d$eta <- with(d, u[f] + 1 + 4 * x)
pz <- 0.2 # probability of excess zeros
zi <- rbinom(100, size = 1, prob = pz)
d$y <- ifelse(zi, 0, rpois(100, lambda = exp(d$eta)))
d$zi <- zi
```

In the next plot, the open circles are excess zeros and the closed circles are the data derived from the Poisson.

```{r}
ggplot(d, aes(x, y, colour = f, shape = as.factor(zi))) + geom_point() +
  scale_shape_manual(values = c(20, 21))
```

Here I have split them up into 2 panels:

```{r}
ggplot(d, aes(x, y, colour = f)) + geom_point() +
  facet_wrap(~zi)
```

The glmmADMB package lets you fit zero-inflated GLMMs, but we are going to use glmmTMB, which is newer, usually faster, and less likely to return strange errors.

```{r, eval=FALSE}
# zip_admb <- glmmADMB::glmmadmb(y ~ x + (1 | f), 
#   data = d, family = "poisson", zeroInflation = TRUE)
# summary(zip_admb)
```

```{r, eval=FALSE}
library(glmmTMB)
zip_tmb <- glmmTMB(y ~ x + (1 | f), data = d, family = poisson(link = "log"), 
  ziformula = ~ 1)
summary(zip_tmb)
```

glmmTMB returns the zero inflation parameter on the logit scale. But we can back transform that with the inverse logit transformation. Remember, the true value was 0.2.

```{r, eval=FALSE}
zi <- coef(summary(zip_tmb))$zi[1, "Estimate"]
zi_se <- coef(summary(zip_tmb))$zi[1, "Std. Error"]

# CIs:
inverse_logit(zi - 1.96 * zi_se) %>% round(2)
inverse_logit(zi) %>% round(2)
inverse_logit(zi + 1.96 * zi_se) %>% round(2)
```

# Hurdle/delta model

An alternative is to fit two separate models: one to the 0s vs non-0s, and another to the positive values only with a truncated distribution.

Run the following code to generate a small simulated data set. 

```{r}
# simulated zero-inflated Poisson example
set.seed(99)
d <- data.frame(f = factor(rep(letters[1:10], each = 10)), x = runif(400))
u <- rnorm(10, sd = 2)
d$eta0 <- with(d, u[f] + 1 + 0.6 * x)
d$eta <- with(d, u[f] + 1 + 4 * x)
pz <- inverse_logit(d$eta0) # probability of excess zeros
zi <- rbinom(400, size = 1, prob = pz)
rtpois <- function(N, lambda) qpois(runif(N, dpois(0, lambda), 1), lambda) # truncated poisson
d$y <- ifelse(zi==0, 0, rtpois(400, lambda = exp(d$eta)))
d$zi <- zi
```

Let's plot the data we just created. Now we have zeros and positive discrete values. 

```{r}
ggplot(d, aes(x, y, colour = f, shape = as.factor(zi))) + geom_point() +
  scale_shape_manual(values = c(21, 20))

ggplot(d, aes(x, y, colour = f)) + geom_point() +
  facet_wrap(~zi)
```

Let's fit 2 separate models. First a truncated Poisson to the positive values. This is a Poisson distribution that starts at one instead of zero.

```{r, eval=FALSE}
hurd_a <- glmmTMB(y ~ x + (1 | f), data = filter(d, y > 0), 
  family = list(family = "truncated_poisson", link = "log"))
summary(hurd_a)
```

Next we fit a binomial model to the 0s vs. non-0s.

```{r, eval=FALSE}
d <- mutate(d, y1 = y > 0)
hurd_b <- glmmTMB(y1 ~ x + (1 | f), data = d, 
  family = binomial(link = "logit"))
summary(hurd_b)
```

We can derive the prediction of the hurdle model by multiplying the probability predicted by the binomial model with the prediction from the positive-value model:

```{r, eval=FALSE}
pred_a <- predict(hurd_a, newdata = d)
pred_b <- predict(hurd_b, newdata = d)

head(pred_a)
head(pred_b)
d$pred <- pred_a * pred_b

ggplot(d, aes(y, pred, colour = f, shape = as.factor(zi))) + 
  geom_point() + scale_shape_manual(values = c(21, 20))

ggplot(d, aes(y, pred, colour = f)) + geom_point() +
  facet_wrap(~zi)
```

We can calculate the AIC of the hurdle model simply by adding the AIC of the 2 component models, although we have nothing to compare it to here. 

```{r, eval=FALSE}
AIC(hurd_a) + AIC(hurd_b)
```

# Positive continuous data with zeros 

A relatively common scenario is to have positive continuous data that also has zeros. 

For example, maybe you are calculating density of sea urchins in quadrats, but some of your quadrats are empty. 
Another example is fish trawl survey data where we have a positive continuous measure of catch per unit effort but sometimes a trawl doesn't bring up any of a given species, so we have zeros.

We can apply a similar concept to above where we model the zeros versus non-zeros and the positive values separately. The difference is that you don't have to worry about a truncated distribution for the positive values. You could use a Gamma distribution for the positive values or use a linear model on a log scale. Again, you can combine your predictions at the end as we did above.

I have written more about this here <http://seananderson.ca/2014/05/18/gamma-hurdle.html>

**Update 2018-02-03: see the delta-Gamma lesson.**

# Additional information 

An excellent write up of a zero-inflation model fit to real data:
<https://groups.nceas.ucsb.edu/non-linear-modeling/projects/owls/WRITEUP/owls.pdf>

The best textbook on the subject for R:
Zuur, A. F. S., A. A. Ieno. 2012. Zero Inflated Models and Generalized Linear Mixed Models with R.

Also the newer:
Zuur A.F.,  Ieno E. N. 2016. A Beginner's Guide to Zero-Inflated Models with R.

The pscl package for fitting zero-inflated models without random effects:
https://CRAN.R-project.org/package=pscl 

The vignette for the pscl R package:
https://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf
